---
title: "Summarize river lengths against abundance ranks and protection status"  
author: "Yi Han"  
date: "2021-11-22"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: The following code analyzes porpoise distribution range covered 
by FPAs (Section 3.1 in the main text) and plots the results as shown in Figure 3.

```{r include=FALSE}
library(tidyverse)
```

## 1. Data analysis  

__Summarize the river lengths by year, abundance ranks.__   
```{r message = FALSE}
data_abund_ranks <- read.csv(file="../../data/analysis_outputs/animal_distribution/segment_wise_porpoise_abundance.csv")

str(data_abund_ranks)
```


```{r}
data_abund_ranks <- data_abund_ranks %>% 
  mutate(f_year = as.factor(f_year)) %>% 
  mutate(abund_ranks = as.factor(abund_ranks))
```

```{r}
 data_abund_ranks %>% 
    group_by(abund_ranks, f_year) %>%
    summarise(length = n()*10)
```

__Summarize the river lengths by year, abundance ranks, and protected status.__   

```{r message = FALSE}
data_abund_ranks %>% 
  group_by(abund_ranks, f_year, reserve) %>%
  summarise(length = n()*10)

data_abund_ranks %>% 
  group_by(abund_ranks, f_year, reserve_without_nj) %>%
  summarise(length = n()*10)
```

## 2. Plot the summarized data  

A quick plot of the river lengths by year, abundance ranks, and protected status.  

```{r}
# ggplot(data_abund_ranks, aes(x = f_year, group=reserve, fill=reserve)) +
#     geom_bar(stat = "count", position = position_stack()) +
#     facet_wrap(~abund_ranks, ncol=5)
```

The following code generates Figure 3 in the main text. Note the "17Y" (i.e., 
reserve in 2017) for the high porpoise abundance rank contains Nanjing reserve, 
which includes five segments (i.e., 50 km). In Figure 3 in the main text, 
we further processed the figure and highlighted these five segments.  

```{r}
# For better plotting, assigning new categories to the dataframe.
data_abund_ranks_plot <- data_abund_ranks %>%
    mutate(year_reserve = case_when(f_year=="2006" & reserve_without_nj=="N" ~ "06N",
                                    f_year=="2006" & reserve_without_nj=="Y" ~ "06Y",
                                    f_year=="2017" & reserve=="N" ~ "17N",
                                    f_year=="2017" & reserve=="Y" ~ "17Y")) %>%
    mutate(year_reserve = factor(year_reserve, levels = c("06N", "06Y", "17N", "17Y"))) %>%
    group_by(f_year, year_reserve, abund_ranks) %>%
    summarise(length_km = n() * 10, .groups = 'drop') 

data_abund_ranks_plot
```

```{r warning = FALSE}
abund_ranks_plot <- 
  ggplot(data_abund_ranks_plot, aes(x = f_year, group = year_reserve, 
                                  color = year_reserve, fill = year_reserve)) +
    geom_bar(aes(y = length_km), stat = "identity", width = 1, position = position_stack(), color = "black") +
    facet_grid(~abund_ranks, switch = "x", labeller = labeller(abund_ranks = c(
        "[0,0.05)" = "Extremely Low [0, 0.05)",
        "[0.05,0.1)" = "Low [0.05, 0.1)",
        "[0.1,0.2)" = "Medium [0.1, 0.2)",
        "[0.2,1)" = "High [0.2, 1)"))) +
    scale_fill_manual(values = c("white", "grey95", "grey58", "grey50")) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 650), breaks = seq(0, 650, 100)) +
    scale_x_discrete(limits = c("trt99", "2006", "2017", "trt99"),
                     breaks = c(NA, "2006", "2017", NA),
                     labels = c("", "2006", "2017", "")) +
    labs(x = "Abundance ranks",
         y = "Length (km)") +
    theme_classic(base_size = 14) +
    theme(
          # strip.text = element_blank(),
          strip.placement = "outside",
          panel.spacing = grid::unit(-2, "lines"),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          legend.position = c(1, 1.05),
          legend.justification = c("right", "top"),
          legend.box.just = "right",
          strip.background = element_blank()) 

abund_ranks_plot
```

```{r}
# ggsave("../../results/figures/fig_3_abundance_ranks.png", abund_ranks_plot, width = 8, height = 5)
```

