---
title: "Examining changes of the segment-wise urban%"
author: "Yi Han"  
date: "2021-11-22"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: The file examines the changes of the segment-wise urban% 
from 2001 to 2017. We conducted cluster analysis and classified river segments 
into four different urban development levels. The results are shown in Table S2.  

```{r include=FALSE}
library(tidyverse)
```

## 1. Data preparation  

Get a dataframe with:  
(1) the initial urban% (i.e., urban% of the year 2001),  
(2) the increment of the urban% from 2001 to 2017, and  
(3) the percentage increase of the urban% from 2001 to 2017.  

```{r}
seg_urban_df <- read.csv(file="../../data/processed/seg_wise_urban.csv")

groups_levels <- seg_urban_df %>% filter(year==2001) %>% pull(groups)
seg_urban_df <- seg_urban_df %>%
    mutate(f_year = as.factor(f_year)) %>%
    mutate(groups = factor(groups, levels=groups_levels)) %>% # get the group levels right
    mutate(f_group = as.numeric(groups))
str(seg_urban_df)
```

```{r}
seg_urban_01_df <- seg_urban_df %>% 
    filter(year==2001) %>% 
    select(groups, f_group, seg_urban_perc) %>%
    rename(seg_urban_perc_01 = seg_urban_perc)
# head(seg_urban_01_df, 5)

seg_urban_17_df <- seg_urban_df %>% 
    filter(year==2017) %>% 
    select(groups, f_group, seg_urban_perc) %>%
    rename(seg_urban_perc_17 = seg_urban_perc)
# head(seg_urban_17_df, 5)

df_urban_incre <- seg_urban_01_df %>%
    left_join(seg_urban_17_df, by = c("groups", "f_group")) %>%
    mutate(increment = seg_urban_perc_17 - seg_urban_perc_01) %>%
    mutate(incre_perc = increment/seg_urban_perc_17) %>%
    mutate(incre_perc = case_when(is.na(incre_perc) ~ 0,
                                         TRUE ~ incre_perc)) %>%
    rename(intl = seg_urban_perc_01, 
           end = seg_urban_perc_17)
head(df_urban_incre)
```

__Plot and check the distribution of the incre_perc.__
```{r}
boxplot(df_urban_incre$incre_perc)
```
## 2. Cluster analysis of the urban development levels  

__Choose and scale the variables used for cluster analysis.__  

```{r}
clust_var <- c("intl", "increment", "incre_perc")

urb_std <- df_urban_incre %>%
  dplyr::select(all_of(clust_var)) %>%
  scale(.) # use scales to standardize all variables
```

__K-mean cluster analysis__   

Identify the optimal number of clusters. Plot the total within-cluster 
sum of squares against each number of clusters k (Figure S2).

```{r}
# K-mean cluster
wssplot <- function(data, nc=15, seed=178){
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")} 

wssplot(urb_std, nc=8) 
```

```{r}
# # Open a PNG device
# png("../../results/figures/fig_S_2_wssplot.png", width = 1600, height = 1200, res = 300)
# 
# # Create the plot
# p <- wssplot(urb_std, nc = 8)
# 
# # Close the device
# dev.off()
```


Based on the following graph and considering the relatively small number of segments, 
we classified segments into four clusters.   

```{r}
set.seed(178) # k-means clustering algorithm starts with k randomly selected centroids,
# using set.seed() function for reproducible results.
k_means_fit <- kmeans(urb_std, 4) 
df_urban_incre_kmean <- df_urban_incre %>%
    mutate(development_levels = factor(k_means_fit$cluster, 
        labels = c("less-developed", "fast-developing", "highly-developed", "developed") )) %>%
    select(groups, intl, increment, incre_perc, development_levels)
head(df_urban_incre_kmean, 5)
tail(df_urban_incre_kmean, 5)
str(df_urban_incre_kmean)
```

__Save the results as a csv file (see Table S2)__

```{r}
# write.csv(df_urban_incre_kmean, 
#           file = "../../data/analysis_outputs/trend_analysis/urban_cluster_results.csv", 
#           row.names = FALSE)
```

