---
title: "Examining changing trends of the segment-wise floodplain area and cropland%"
author: "Yi Han"  
date: "2021-11-20"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: The file examines the changes of the floodplain area and the 
cropland% for each of segments from 2001 to 2016. The results are shown in Table S1.  

```{r include=FALSE}
library(tidyverse)
```

## 1. Prepare the data  

__Get the segment-wise floodplain area and cropland% data ready.__  

We excluded the last 50 km-long river section, i.e., the last 5 segments within 
[1450 km, 1500 km) from the following trend analyses because this region is 
close to the estuary, where tidal effects also contribute substantially to 
the variation of the floodplain area.  

```{r}
seg_floodplain_cropland_df <- read.csv(file="../../data/processed/seg_wise_floodplain_ndvi.csv")
str(seg_floodplain_cropland_df)
```

```{r, message=FALSE}
groups_levels <- seg_floodplain_cropland_df %>% filter(year==2003) %>% pull(groups)
seg_floodplain_cropland_df <- seg_floodplain_cropland_df %>%
    mutate(f_year = as.factor(f_year)) %>%
    mutate(groups = factor(groups, levels=groups_levels)) %>% # get the group levels right
    filter(f_group <= 145) %>%
    droplevels()
str(seg_floodplain_cropland_df)
```

## 1. Linear regression of the floodplain area  

__Linear regression of the floodplain area for each segment.__  

```{r}
len <- length(levels(seg_floodplain_cropland_df$groups))
coef_floodplain_trend_list <- list()

for (i in 1:len) {
    # i=27
    groupi <- levels(seg_floodplain_cropland_df$groups)[i]
    
    segment_i <- seg_floodplain_cropland_df %>%
        dplyr::filter(groups %in% groupi) 
    
    lin_modi <- lm(seg_floodplain_area * 10 ~ year, data = segment_i)
        # seg_floodplain_area is the flooplain area per 1 km for the segment
        # thus, to get the total area of the floodplain need to mltiply 
        # seg_floodplain_area by 10.
    
    p_vali <- summary(lin_modi)$coefficients[2, 4] %>% round(., digits = 3)
    r_squaredi <- summary(lin_modi)$r.squared %>% round(., digits = 3)
    slopei <- summary(lin_modi)$coefficients[2, 1] %>% round(., digits = 3)
    
    if(p_vali <= 0.01){
        p_catei <- "<0.01"
      
    } else if(p_vali > 0.01 & p_vali <=0.05){
        p_catei <- "<0.05"
  
    } else {
        p_catei <- as.character(round(p_vali, digit = 3))
    }
    
    coef_floodplain_trend_list[[i]] <- c(groupi, p_vali, p_catei, r_squaredi, slopei)
}

# coef_floodplain_trend_list
```

__Save the results of the lienar regression as a dataframe.__  

```{r}
coef_floodplain_df <- do.call(rbind.data.frame, coef_floodplain_trend_list)
colnames(coef_floodplain_df) <- c("groups", "p_value", "p_category", "r_squared", "slope")

coef_floodplain_df <- coef_floodplain_df %>%
    dplyr::mutate(p_value = as.numeric(as.character(p_value)),
                  r_squared = as.numeric(as.character(r_squared)),
                  slope = as.numeric(as.character(slope)))
head(coef_floodplain_df, 5)
```

__Save the dataframe as a csv file. (Table S1)__
```{r}
# write.csv(df_urban_incre_kmean, 
#           file = "../../data/analysis_outputs/trend_analysis/floodplain_area_regression_results.csv", 
#           row.names = FALSE)
```

## 3. Linear regression of the cropland%  

__Linear regression of the cropland% for each segment.__   

Cropland%, as proportions (range from 0 to 1), is arcsine transformed in linear regression.

```{r}
len <- length(levels(seg_floodplain_cropland_df$groups))
coef_crop_perc_list <- list()

for (i in 1:len) {
    # i=19
    groupi <- levels(seg_floodplain_cropland_df$groups)[i]
    
    segment_i <- seg_floodplain_cropland_df %>%
        dplyr::filter(groups %in% groupi) 
    
    segment_i$asin_cropland_perc <- asin(sqrt(segment_i$seg_cropland_perc))
    crop_perc_modi <- lm(asin_cropland_perc ~ year, data=segment_i)
    
    p_vali <- summary(crop_perc_modi)$coefficients[2, 4] %>% round(., digits = 3)
    r_squaredi <- summary(crop_perc_modi)$r.squared %>% round(., digits = 3)
    slopei <- summary(crop_perc_modi)$coefficients[2, 1] %>% round(., digits = 3)
    
    if(p_vali <= 0.01){
        p_catei <- "<0.01"
      
    } else if(p_vali > 0.01 & p_vali <=0.05){
        p_catei <- "<0.05"
      
    } else {
        p_catei <- as.character(round(p_vali, digit = 3))
    }
    
    coef_crop_perc_list[[i]] <- c(groupi, p_vali, p_catei, r_squaredi, slopei)
}
```

__Save the results of the linear regression as a dataframe and a csv file (Table S1).__  

```{r}
coef_crop_perc_df <- do.call(rbind.data.frame, coef_crop_perc_list)
colnames(coef_crop_perc_df) <- c("groups", "p_value", "p_cate", "r_squared", "slope")
coef_crop_perc_df <- coef_crop_perc_df %>%
    dplyr::mutate(p_value = as.numeric(as.character(p_value)),
                  r_squared = as.numeric(as.character(r_squared)),
                  slope = as.numeric(as.character(slope)))
str(coef_crop_perc_df)
head(coef_crop_perc_df, 5)
```

__Save the results as a csv file (see Table S1)__

```{r}
# write.csv(df_urban_incre_kmean, 
#           file = "../../data/analysis_outputs/trend_analysis/cropland_perc_regression_results.csv", 
#           row.names = FALSE)
```


