---
title: "Analyzing initial habitat conditions between PAs and non-PAs"
author: "Yi Han"  
date: "2021-11-25"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: This file tests whether PAs had more river segments with 
stable habitat conditions than the non-PAs. It generates statistical outputs 
and produces Fig.5d-f in the main text.   

It uses data in the files 'floodplain_area_regression_results.csv', 
'cropland_perc_regression_results.csv', 'urban_cluster_results.csv' 
in the 'trend_analysis_habitats' folder.  

```{r include=FALSE}
library(tidyverse)
# library(fifer)
```

## 1. Examine if the PAs hold more river segments with stable floodplain area   

__Prepare the dataframe:__ we grouped river segments into decreased and stable 
categories given whether the floodplain area showed a significant decline. 
We grouped the nine segments with a decreased floodplain area into the stable 
category (see Appendix S3 for more details).  

```{r}
seg_wise_floodplain_df <- read.csv(file="../trend_analysis_habitats/floodplain_area_regression_results.csv")
group_lvls <- seg_wise_floodplain_df$groups
floodplain_df <- seg_wise_floodplain_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    mutate(flp_area_trend = case_when((p_value < 0.05 & slope < 0) ~ "decreased",
                                TRUE ~ "stable")) %>%
    mutate(flp_area_trend = factor(flp_area_trend, 
                                    levels = c("stable", "decreased"))) %>%
    mutate(reserve_without_nj = case_when(
        group_levels %in% c(22:30, 41:53, 88:104, 110:115, 139:141) ~ "Y",
        TRUE ~ "N")) %>%
    mutate(reserve_without_nj = as.factor(reserve_without_nj)) 
head(floodplain_df, 5)
```

__Plot the number of river segments of the PA and non-PA groups at each category__ 
__of the floodplain area trend.__ (Fig. 5d)   

```{r}
ggplot(floodplain_df, aes(x = reserve_without_nj, 
                          color = flp_area_trend, fill=flp_area_trend)) +
      geom_bar(stat = "count", width=0.6, position = position_dodge(), color="black") + 
      scale_fill_manual(values = c("white", "grey50")) +
      scale_y_continuous(expand = c(0, 0), limits = c(0, 80), breaks=seq(0, 80, 10)) +
      scale_x_discrete(limits = c("Y", "N"),
                       breaks = c("Y", "N"),
                       labels = c("PAs", "non-PAs")) +
      labs (x = "", y = "Segment counts") +
      theme_classic(base_size = 14) +
      theme(legend.position = c(.1, .99),
            legend.justification = c("left", "top"),
            legend.box.just = "left")
```
__Chi-square test__   

```{r}
tbl_flp_trend <- table(floodplain_df$flp_area_trend, floodplain_df$reserve_without_nj)
tbl_flp_trend
chisq.test(tbl_flp_trend)
```
The stable group actually contains 9 segments with increased floodplain areas, where
8 from non-PAs and 1 from PAs. We now remove them for chi square analysis.

```{r}
tbl_flp_trend_remove_inc <- as.table(rbind(c(66, 39), c(23, 8)))
dimnames(tbl_flp_trend_remove_inc) <- list(group = c("stable", "decreased"),
                                           PA = c("N", "Y"))
tbl_flp_trend_remove_inc
chisq.test(tbl_flp_trend_remove_inc)
```

## 2. Examine if the PAs hold more river segments with stable cropland%   

__Prepare the dataframe:__ we grouped river segments into increased and stable 
categories given whether the cropland% showed a significant increase. We grouped 
the four segments with an increased cropland% into the stable category 
(see Appendix S3 for more details).  

```{r}
seg_wise_cropland_df <- read.csv(file="../trend_analysis_habitats/cropland_perc_regression_results.csv")
cropland_df <- seg_wise_cropland_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    mutate(crop_perc_trend = case_when((p_value < 0.05 & slope > 0) ~ "increased",
                             TRUE ~ "stable")) %>% 
    mutate(crop_perc_trend = factor(crop_perc_trend, 
                                    levels = c("stable", "increased"))) %>%
    mutate(reserve_without_nj = case_when(
        group_levels %in% c(22:30, 41:53, 88:104, 110:115, 139:141) ~ "Y",
        TRUE ~ "N")) %>%
    mutate(reserve_without_nj = as.factor(reserve_without_nj)) 
head(cropland_df, 5)
```

__Plot the number of river segments of the PA and non-PA groups at each category__ 
__of the cropland% trend.__ (Fig. 5e)   

```{r}
ggplot(cropland_df, aes(x=reserve_without_nj, color=crop_perc_trend, fill=crop_perc_trend)) +
      geom_bar(stat = "count", width=0.6, position = position_dodge(), color="black") + 
      scale_fill_manual(values = c("white", "grey50")) +
      scale_y_continuous(expand = c(0, 0), limits = c(0, 80), breaks=seq(0, 80, 10)) +
      scale_x_discrete(limits = c("Y", "N"),
                       breaks = c("Y", "N"),
                       labels = c("PAs", "non-PAs")) +
      labs (x = "", y = "Segment counts") +
      theme_classic(base_size = 14) +
      theme(legend.position = c(.1, .99),
            legend.justification = c("left", "top"),
            legend.box.just = "left")
```

__Chi-square test__   

```{r}
tbl_crop_trend <- table(cropland_df$crop_perc_trend, cropland_df$reserve_without_nj)
tbl_crop_trend
chisq.test(tbl_crop_trend)
```

The stable group actually contains 4 segments with decreased cropland%, where
all 4 from non-PAs. We now remove them for Chi square analysis.

```{r}
tbl_crop_trend_remove_dec <- as.table(rbind(c(62, 24), c(31, 24)))
dimnames(tbl_crop_trend_remove_dec) <- list(group = c("stable", "increased"),
                                           PA = c("N", "Y"))
tbl_crop_trend_remove_dec 
chisq.test(tbl_crop_trend_remove_dec)
```

## 3. Examine if the PAs hold more river segments with lower urban development level   

__Prepare the dataframe:__ we grouped the “developed” and the “highly developed” 
as one category as “developed”, representing those river segments with relatively 
large urban% and large increment in urban% (see Appendix S3 for more details).   

```{r}
seg_wise_urban_df <- read.csv(file="../trend_analysis_habitats/urban_cluster_results.csv")

urban_df <- seg_wise_urban_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    droplevels() %>%
    mutate(urban_levels = case_when(
       development_levels=="less-developed" ~ "Less developed",
       development_levels=="fast-developing" ~ "Fast developing",
       TRUE ~ "Developed")) %>%
    mutate(urban_levels = factor(urban_levels, 
       levels = c("Less developed", "Fast developing", "Developed"))) %>%
    mutate(reserve_without_nj = case_when(
        group_levels %in% c(22:30, 41:53, 88:104, 110:115, 139:141) ~ "Y",
        TRUE ~ "N")) %>%
    mutate(reserve_without_nj = as.factor(reserve_without_nj)) 
head(urban_df, 5)
```

__Plot the number of river segments of the PA and non-PA groups at each category__ 
__of the urban development levels.__ (Fig. 5f)   

```{r}
ggplot(urban_df, aes(x=reserve_without_nj, color=urban_levels, fill=urban_levels)) + 
    geom_bar(stat = "count", width=0.8, 
             position=position_dodge(preserve = "single"), color="black") +
    scale_fill_manual(values = c("white", "grey80", "grey30")) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 44), breaks=seq(0, 44, 10)) +
    scale_x_discrete(limits = c("Y", "N"),
                     breaks = c("Y", "N"),
                     labels = c("PAs", "non-PAs")) +
    labs (x = "", y = "Segment counts") +
    theme_classic(base_size = 14) +
    theme(legend.position = c(.1, .99),
          legend.justification = c("left", "top"),
          legend.box.just = "left")
```

__Chi-square test__   

```{r}
tbl_urban_trend <- table(urban_df$urban_levels, urban_df$reserve_without_nj)
chisq.test(tbl_urban_trend)
```

__Post hoc pairwise comparisons__   

Use the function `fisher.multcomp` in package `RVAideMemoire`.  

```{r}
library(RVAideMemoire)
RVAideMemoire::fisher.multcomp(tbl_urban_trend, p.method = "bonferroni")
```


