---
title: "Analyzing initial habitat conditions between PAs and non-PAs"
author: "Yi Han"  
date: "2021-11-25"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: The following code aims to test whether the PAs had better 
initial habitat conditions than the non-PAs: including larger floodplain area, and
smaller human disturbances which reflected by smaller cropland% and urban%. The
results of this analysis correspond to Fig 5 a-c.

```{r include=FALSE}
library(tidyverse)
library(gghalves)
```

## 1. Compare initial floodplain area and cropland% between PAs and non-PAs  
We used the mean floodplain area, mean cropland%, and mean urban% of the period 
from 2001 to 2006 to represent the initial floodplain habitat conditions. 

```{r}
seg_floodplain_cropland_df <- read.csv(file="../data_preparation/seg_wise_floodplain_ndvi.csv")
str(seg_floodplain_cropland_df)
```

### 1.1 Prepare the dataframe and add reserve information  
(1) reserve: the segments that are identified as reserves in 2017   
(2) reserve_without_nj: the segments that are identified as reserves before 2006, 
    Nanjing (1261:1308) was excluded as it was established in 2014   
(3) reserve_names: the name of the reserve   

```{r, message=FALSE}
groups_levels <- seg_floodplain_cropland_df %>% filter(year==2003) %>% pull(groups)
floodplain_initial_df <- seg_floodplain_cropland_df %>%
    mutate(f_year = as.factor(f_year),
           groups = factor(groups, levels=groups_levels)) %>% # get the group levels right
    filter(year <= 2006) %>%
    dplyr::group_by(groups, f_group) %>%
    dplyr::summarise(initial_floodplain = mean(seg_floodplain_area),
                     initial_crop = mean(seg_ndvi_5_7),
                     initial_crop_perc = mean(seg_cropland_perc)) %>%
    mutate(reserve = case_when(f_group %in% c(22:30, 41:53, 88:104, 110:115, 127:131, 139:141) ~ "Y",
                               TRUE ~ "N")) %>%
    mutate(reserve_without_nj = case_when(f_group %in% c(22:30, 41:53, 88:104, 110:115, 139:141) ~ "Y",
                                          TRUE ~ "N")) %>%
    mutate(reserve_name = case_when(f_group %in% c(22:30) ~ "shishou",
                                    f_group %in% c(41:53) ~ "honghu",
                                    f_group %in% c(88:104) ~ "anqing",
                                    f_group %in% c(110:115) ~ "tongling",
                                    f_group %in% c(127:131) ~ "nanjing",
                                    f_group %in% c(139:141) ~ "zhenjiang",
                                    TRUE ~ "non-PAs")) %>%
    mutate(reserve = as.factor(reserve),
           reserve_without_nj = as.factor(reserve_without_nj),
           reserve_name = as.factor(reserve_name)) %>%
    ungroup()
str(floodplain_initial_df)
```

### 1.2 Compare the initial floodplain areas between the PAs and non-PAs  
__Plot the initial floodplain areas between the PAs and non-PAs__  
```{r}
ggplot(floodplain_initial_df, aes(x = reserve_without_nj, y = initial_floodplain)) +
    geom_half_boxplot(fill="grey70")+
    geom_half_point(color="grey30")+
    scale_y_continuous(limits = c(0, 0.8), breaks=seq(0, 0.8, 0.1)) +
    scale_x_discrete(limits = c("Y", "N"),
                     breaks = c("Y", "N"),
                     labels = c("PAs", "non-PAs")) +
    labs (x = "", y = expression(paste("Floodplain area (km"^"2", ") km"^" -1"))) +
    theme_classic(base_size = 14) +
    theme(legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")
```
__Compare the initial floodplain areas between the PAs and non-PAs using the Mann–Whitney U test__  
```{r}
floodplain_initial_PAs <- floodplain_initial_df %>% 
    filter(reserve_without_nj=="Y") %>% pull(initial_floodplain)
floodplain_initial_nonPAs <- floodplain_initial_df %>% 
    filter(reserve_without_nj=="N") %>% pull(initial_floodplain)

floodplain_initial_outputs <- c(mean(floodplain_initial_PAs), sd(floodplain_initial_PAs), 
                          mean(floodplain_initial_nonPAs), sd(floodplain_initial_nonPAs))
names(floodplain_initial_outputs) <- c("mean floodplain_initial_PAs", "sd floodplain_initial_PAs",
                                 "mean floodplain_initial_nonPAs", "sd floodplain_initial_nonPAs")
floodplain_initial_outputs

wilcox.test(initial_floodplain ~ reserve_without_nj, data=floodplain_initial_df)
```

### 1.3 Compare the initial cropland% between the PAs and non-PAs 
__Plot the initial cropland% between the PAs and non-PAs__  
```{r}
ggplot(floodplain_initial_df, aes(x=reserve_without_nj, y=initial_crop_perc)) +
    geom_half_boxplot(fill="grey70")+
    geom_half_point(color="grey30")+
    scale_y_continuous(limits = c(0, 0.6), breaks=seq(0, 0.6, 0.2),
                       labels = scales::percent_format(accuracy = 1)) +
    scale_x_discrete(limits = c("Y", "N"),
                     breaks = c("Y", "N"),
                     labels = c("PAs", "non-PAs")) +
    labs (x = "", y = "cropland%") +
    theme_classic(base_size = 14) +
    theme(legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")
```

__Compare the initial cropland% between the PAs and non-PAs using the Mann–Whitney U test__ 
```{r}
crop_perc_initial_PAs <- floodplain_initial_df %>% 
    filter(reserve_without_nj=="Y") %>% pull(initial_crop_perc)
crop_perc_initial_nonPAs <- floodplain_initial_df %>% 
    filter(reserve_without_nj=="N") %>% pull(initial_crop_perc)

crop_perc_initial_outputs <- c(mean(crop_perc_initial_PAs), sd(crop_perc_initial_PAs), 
                          mean(crop_perc_initial_nonPAs), sd(crop_perc_initial_nonPAs))
names(crop_perc_initial_outputs) <- c("mean crop_perc_initial_PAs", "sd crop_perc_initial_PAs",
                                 "mean crop_perc_initial_nonPAs", "sd crop_perc_initial_nonPAs")
crop_perc_initial_outputs

wilcox.test(initial_crop_perc ~ reserve_without_nj, data=floodplain_initial_df)
```

## 2. Compare initial urban% between PAs and non-PAs  

```{r}
seg_urban_df <- read.csv(file="../data_preparation/seg_wise_urban.csv")
str(seg_urban_df)
```

### 2.1 Prepare the dataframe and add reserve information  
```{r}
urban_initial_df <- seg_urban_df %>% 
    mutate(f_year = as.factor(f_year),
           groups = factor(groups, levels=groups_levels)) %>% # get the group levels right
    filter(year <= 2006) %>%
    group_by(groups) %>%
    summarise(urban_perc_initial = mean(seg_urban_perc)) %>%
    mutate(f_group = c(1:150)) %>%
    mutate(reserve = case_when(f_group %in% c(22:30, 41:53, 88:104, 110:115, 127:131, 139:141) ~ "Y",
                               TRUE ~ "N")) %>%
    mutate(reserve_without_nj = case_when(f_group %in% c(22:30, 41:53, 88:104, 110:115, 139:141) ~ "Y",
                                          TRUE ~ "N")) %>%
    mutate(reserve_name = case_when(f_group %in% c(22:30) ~ "shishou",
                                    f_group %in% c(41:53) ~ "honghu",
                                    f_group %in% c(88:104) ~ "anqing",
                                    f_group %in% c(110:115) ~ "tongling",
                                    f_group %in% c(127:131) ~ "nanjing",
                                    f_group %in% c(139:141) ~ "zhenjiang",
                                    TRUE ~ "non-PAs")) %>%
    mutate(reserve = as.factor(reserve),
           reserve_without_nj = as.factor(reserve_without_nj),
           reserve_name = as.factor(reserve_name)) %>%
    # filter(f_group <= 145) %>%
    droplevels() %>%
    ungroup()
summary(urban_initial_df)
```

__Plot the initial urban% between the PAs and non-PAs__   
```{r}
ggplot(urban_initial_df, aes(x = reserve_without_nj, y = urban_perc_initial)) +
  geom_half_boxplot(fill="grey70") +
  geom_half_point(color="grey30") +
  scale_y_continuous(limits = c(0, 1), breaks=seq(0, 1, 0.2),
                     labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(limits = c("Y", "N"),
                   breaks = c("Y", "N"),
                   labels = c("PAs", "non-PAs")) +
  labs (x = "", y = "Urban%") +
  theme_classic(base_size = 14) +
  theme(legend.position = c(.99, .99),
        legend.justification = c("right", "top"),
        legend.box.just = "right")
```

__Compare the initial urban% between the PAs and non-PAs using the Mann–Whitney U test__   
```{r}
urban_inital_PAs <- urban_initial_df %>% filter(reserve_without_nj =="Y") %>% pull(urban_perc_initial)
urban_inital_nonPAs <- urban_initial_df %>% filter(reserve_without_nj =="N") %>% pull(urban_perc_initial)

urban_inital_outputs <- c(mean(urban_inital_PAs), sd(urban_inital_PAs), 
                          mean(urban_inital_nonPAs), sd(urban_inital_nonPAs))

names(urban_inital_outputs) <- c("mean urb_inital_PAs", "sd urb_inital_PAs",
                                 "mean urb_inital_nonPAs", "sd mean urb_inital_nonPAs")
urban_inital_outputs

wilcox.test(urban_perc_initial ~ reserve_without_nj, data=urban_initial_df)
```

