---
title: "Summarizing features of representative river sections of porpoise distribution"
author: "Yi Han"  
date: "2021-12-03"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: The following code first summarized the segment-wise porpoise 
abundance changes, changing trends of the floodplain area and cropland%, 
and urban development levels (see section 1. segment-wise porpoise abundance changes 
and habitat changes). This piece of code was derived from the file
_'plot_segment_wise_habitat_changes_and_porpoise_gaps.Rmd'_ in folder
_'trend_analysis_habitats\'_. 

```{r include=FALSE}
library(tidyverse)
library(mclust)
library(nnet)
library(broom)
library(gridExtra)
```

## 1. segment-wise porpoise abundance changes and habitat changes

1.1 We define a distribution gap of the porpoise as a river sections with extremely
low porpoise abundance, i.e., < 0.05 porpoise per km, in both 2006 and 2017.   

```{r}
seg_wise_abund_df <- read.csv(file="../02_percentage_coverage_of_PAs/segment_wise_porpoise_abundance.csv")
head(seg_wise_abund_df, 5)
```

```{r}
group_levels <- seg_wise_abund_df$groups[1:150]
abund_df <- seg_wise_abund_df %>%
    dplyr::select(groups, year, segment_mean, reserve, reserve_without_nj) %>%
    dplyr::mutate(groups = factor(groups, levels=group_levels)) %>%
    dplyr::mutate(group_levels = as.numeric(groups)) %>%
    tidyr::pivot_wider(., names_from = year, values_from = segment_mean) %>%
    dplyr::rename(abund_06 = "2006", abund_17 = "2017") %>%
    dplyr::mutate(abund_change = abund_17 - abund_06) %>%
    dplyr::filter(group_levels <= 145) |> 
    dplyr::select(groups, group_levels, abund_06, abund_17, abund_change, reserve, reserve_without_nj)
head(abund_df, 5)
```

1.2 We grouped river segments into decreased and stable categories given whether 
the floodplain area showed a significant decline. We grouped the nine segments 
with a decreased floodplain area into the stable category (see Appendix S3 for 
more details).  

```{r}
seg_wise_floodplain_df <-
    read.csv(file="../trend_analysis_habitats/floodplain_area_regression_results.csv")

group_lvls <- seg_wise_floodplain_df$groups
floodplain_df <- seg_wise_floodplain_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    mutate(flp_area_trend = case_when((p_value < 0.05 & slope < 0) ~ "decreased",
                                TRUE ~ "stable")) %>%
    mutate(flp_area_trend = factor(flp_area_trend, 
                                    levels = c("decreased", "stable"))) |> 
    dplyr::select(groups, group_levels, flp_area_trend)

head(floodplain_df, 5)
```

1.3 We grouped river segments into increased and stable categories given whether 
the cropland% showed a significant increase. We grouped the four segments 
with an increased cropland% into the stable category (see Appendix S3 for 
more details).  

```{r}
seg_wise_cropland_df <- read.csv(file="../trend_analysis_habitats/cropland_perc_regression_results.csv")

cropland_df <- seg_wise_cropland_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    mutate(crop_perc_trend = case_when((p_value < 0.05 & slope > 0) ~ "increased",
                             TRUE ~ "stable")) %>% 
    mutate(crop_perc_trend = factor(crop_perc_trend, 
                                    levels = c("increased", "stable"))) |> 
    dplyr::select(groups, group_levels, crop_perc_trend)
head(cropland_df, 5)
```

We grouped the “developed” and the “highly developed” as one category as 
“developed”, representing those river segments with relatively large urban% 
and large increment in urban% (see Appendix S3 for more details).  

```{r}
seg_wise_urban_df <- read.csv(file="../trend_analysis_habitats/urban_cluster_results.csv")

urban_df <- seg_wise_urban_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    filter(group_levels <= 145) %>%
    droplevels() %>%
    mutate(urban_levels = case_when(
       development_levels=="less-developed" ~ "Less developed",
       development_levels=="fast-developing" ~ "Fast developing",
       TRUE ~ "Developed")) %>%
    mutate(urban_levels = factor(urban_levels, 
       levels = c("Developed", "Fast developing", "Less developed"))) |> 
    dplyr::select(groups, group_levels, urban_levels)
head(urban_df, 5)
```

Combine above dataframes into a single dataframe for multivariate linear regression

```{r}
df <- abund_df |> 
    left_join(floodplain_df, by=c("groups", "group_levels")) |> 
    left_join(cropland_df, by=c("groups", "group_levels")) |> 
    left_join(urban_df, by=c("groups", "group_levels"))
df
```

## 2. Clustering Porpoise Abundance Changing Patterns using Gaussiam Mixture Model

```{r}
data_scaled <- scale(df[, c("abund_06", "abund_17", "abund_change")])
```


Try models with a range of Gaussian components and covariance structures

```{r}
numComponents <- 1:5
covarianceModels <- c("EII", "EEI", "VII", "VEE", "EVI", "EEE", "EVE", "EVV", "VVV")

# Initialize a data frame to store BIC values for each model
bic_values <- matrix(nrow = length(numComponents), ncol = length(covarianceModels), dimnames = list(numComponents, covarianceModels))

# Loop through each type of covariance model
for(model in covarianceModels) {
    # Fit the GMM for each covariance type across a range of components
    fit <- Mclust(data_scaled, G = numComponents, modelNames = model)
    print(model)
    print(fit$BIC)
}
```

# Find the model and number of components with the highest BIC

```{r}
gmm_result <- Mclust(data_scaled, G=5, modelNames = "VII")
print(gmm_result$BIC)
summary(gmm_result)
df$gmm_cluster <- as.factor(gmm_result$classification)
```

```{r}
pattern_plot1 <- ggplot(
  df, aes(x = abund_06, y = abund_17, size = abund_change, color = gmm_cluster)
  ) +
  geom_point(alpha = 0.8) +  
  scale_size_continuous(range = c(0.1, 3)) +
  labs(x = expression(paste("Encounter rate in 2006 ", "(", "porpoise km"^" -1", ")")),
       y = expression(paste("Encounter rate in 2017 ", "(", "porpoise km"^" -1", ")")),
       size = expression(paste("Encounter rate in 2017 ", "(", "porpoise km"^" -1", ")")),
       color = "Porpose abundance changing patterns") +
  # ggtitle("Scatter Plot of Porpoise Abundance by Cluster") +
  theme_minimal()
pattern_plot1
```

```{r}
pattern_plot2 <- ggplot(
  df, aes(x = abund_06, y = abund_change, size = abund_17, color = gmm_cluster)
  ) +
  geom_point(alpha = 0.8) +  
  scale_size_continuous(range = c(0.1, 3)) +
  labs(x = expression(paste("Encounter rate in 2006 ", "(", "porpoise km"^" -1", ")")),
       y = expression(paste("Change in encounter rates ", "(", "porpoise km"^" -1", ")")),
       size = expression(paste("Encounter rate in 2017 ", "(", "porpoise km"^" -1", ")")),
       color = "Porpose abundance changing patterns") +
  # ggtitle("Scatter Plot of Porpoise Abundance by Cluster") +
  theme_minimal()
pattern_plot2
```
```{r}
combined_plot <- grid.arrange(pattern_plot1, pattern_plot2, ncol = 1) 
# combined_plot
```

```{r}
ggsave("combined_plots_abundance_clustesrs.png", combined_plot, width = 8, height = 8)
```

We categorized river segments into five porpoise abundance patterns based on the outputs of the  Gaussian Mixture Model (GMM): 
- no presence
- consistently low
- low-to-medium with minor decline
- medium-to-high with major decline
- increasing or high

## multinominal logistic regression

```{r}
df$gmm_cluster <- factor(df$gmm_cluster, levels = c("2", "1", "3", "5", "4"))

df$gmm_cluster <- fct_recode(df$gmm_cluster,
                           "no_presence" = "2",
                           "low_subtle_increase" = "1",
                           "low_med_minor_decline" = "3",
                           "med_high_major_decline" = "5", 
                           "increasing_or_high" = "4")
print(levels(df$gmm_cluster))
```

```{r}
df$flp_area_trend <- factor(df$flp_area_trend, levels = c("stable", "decreased"))
df$crop_perc_trend <- factor(df$crop_perc_trend, levels = c("stable", "increased"))
df$urban_levels <- factor(df$urban_levels, 
                          levels = c("Less developed", "Fast developing", "Developed"))
```


```{r}
df_summary_agg_vals <- df |>  
  group_by(gmm_cluster) |> 
  summarise(mean_abund_17 = round(mean(abund_17), 3),
            mean_abund_change = round(mean(abund_change), 3))
df_summary_agg_vals
```


```{r}
df$gmm_cluster <- relevel(df$gmm_cluster, ref = "no_presence")

multinom_model <- multinom(gmm_cluster ~ flp_area_trend + crop_perc_trend + urban_levels + reserve, 
                           data = df)

mult_output <- tidy(multinom_model, conf.int = TRUE, exponentiate = TRUE) |> 
    mutate_if(is.numeric, round, 3) |> 
    dplyr::filter(p.value < 0.1)

mult_output
```

```{r}
df$gmm_cluster <- relevel(df$gmm_cluster, ref = "low_subtle_increase")

multinom_model <- multinom(gmm_cluster ~ flp_area_trend + crop_perc_trend + urban_levels + reserve, 
                           data = df)

mult_output <- tidy(multinom_model, conf.int = TRUE, exponentiate = TRUE) |> 
    mutate_if(is.numeric, round, 3) |> 
    dplyr::filter(p.value < 0.1)

mult_output
```
```{r}
df$gmm_cluster <- relevel(df$gmm_cluster, ref = "low_med_minor_decline")

multinom_model <- multinom(gmm_cluster ~ flp_area_trend + crop_perc_trend + urban_levels + reserve, 
                           data = df)

mult_output <- tidy(multinom_model, conf.int = TRUE, exponentiate = TRUE) |> 
    mutate_if(is.numeric, round, 3) |> 
    dplyr::filter(p.value < 0.1)

mult_output
```

```{r}
df$gmm_cluster <- relevel(df$gmm_cluster, ref = "med_high_major_decline")

multinom_model <- multinom(gmm_cluster ~ flp_area_trend + crop_perc_trend + urban_levels + reserve, 
                           data = df)

mult_output <- tidy(multinom_model, conf.int = TRUE, exponentiate = TRUE) |> 
    mutate_if(is.numeric, round, 3) |> 
    dplyr::filter(p.value < 0.1)

mult_output
```

```{r}
df_summary <- df |> 
    dplyr::select(groups, gmm_cluster, flp_area_trend, crop_perc_trend, urban_levels, reserve) |> 
    group_by(gmm_cluster) |> 
    summarise(
        Total_length = n(),
        Decreased_Flp_Area = paste0(round(mean(flp_area_trend == "decreased") * 100, 2), "%"),
        Increased_Crop = paste0(round(mean(crop_perc_trend == "increased") * 100, 2), "%"),
        Less_Developed_Urban = paste0(round(mean(urban_levels == "Less developed") * 100, 2), "%"),
        Fast_Developing_Urban = paste0(round(mean(urban_levels == "Fast developing") * 100, 2), "%"),
        Developed_Urban = paste0(round(mean(urban_levels == "Developed") * 100, 2), "%"),
        Reserve_Y = paste0(round(mean(reserve == "Y") * 100, 2), "%"),
        Groups_String = toString(groups),
        # Groups_List = list(groups)
      )
```

```{r}
order <- c("no_presence", "low_subtle_increase", "low_med_minor_decline", "med_high_major_decline", "increasing_or_high")
df_summary$gmm_cluster <- factor(df_summary$gmm_cluster, levels = order)
df_summary <- df_summary[order(df_summary$gmm_cluster), ]
df_summary

df_summary_agg_vals$gmm_cluster <- factor(df_summary_agg_vals$gmm_cluster, levels = order)
df_summary_agg_vals <- df_summary_agg_vals[order(df_summary_agg_vals$gmm_cluster), ]
df_summary_agg_vals

df_summary_sim <- df_summary |> 
  left_join(df_summary_agg_vals, by = "gmm_cluster") |> 
  select(!Groups_String)
df_summary_sim
```

```{r}
df_summary_full <- df_summary |> 
  left_join(df_summary_agg_vals, by = "gmm_cluster")
df_summary_full$Groups_String
```