---
title: "Plot segment-wise habitat changes and porpoise distribution gaps"
author: "Yi Han"  
date: "2021-11-26"
output: html_notebook
---
**Email**: yi.han.178@gmail.com  
**Description**: Plot the segment-wise porpoise abundance changes, changing 
trends of the floodplain area and cropland%, and urban development levels, which
correspond to Fig. 2b.

```{r include=FALSE}
library(tidyverse)
```

## 1. Plot distribution gaps of the porpoise    

We define a distribution gap of the porpoise as a river sections with extremely
low porpoise abundance, i.e., < 0.05 porpoise per km, in both 2006 and 2017.   

```{r}
seg_wise_abund_df <- read.csv(file="../02_percentage_coverage_of_PAs/segment_wise_porpoise_abundance.csv")
head(seg_wise_abund_df, 5)
```

```{r}
group_levels <- seg_wise_abund_df$groups[1:150]
abund_df <- seg_wise_abund_df %>%
    dplyr::select(groups, year, segment_mean) %>%
    dplyr::mutate(groups = factor(groups, levels=group_levels)) %>%
    dplyr::mutate(group_levels = as.numeric(groups)) %>%
    tidyr::pivot_wider(., names_from = year, values_from = segment_mean) %>%
    dplyr::rename(abund_06 = "2006", abund_17 = "2017") %>%
    dplyr::mutate(abund_change = abund_17 - abund_06) %>%
    dplyr::filter(group_levels <= 145)
head(abund_df, 5)
```

```{r}
abund_df <- abund_df %>%
    mutate(dist_gap = case_when((abund_06 < 0.05 & abund_17 < 0.05) ~ "perm_gap", 
                                (abund_06 >= 0.05 & abund_17 < 0.05) ~ "new_gap",
                                (abund_06 <= abund_17 & abund_17 >= 0.2) ~ "steady-recover",
                                (abund_06 >= 0.2 & abund_17 >= 0.2) ~ "consist high",
                                TRUE ~ "non-gap-non-high")) %>%
    mutate(dist_gap = factor(dist_gap, levels = c("perm_gap", "new_gap", "non-gap-non-high", "consist high", "steady-recover")))
head(abund_df, 5)
```

__Plot the distribution gaps__  

```{r} 
abund_gap_plot <- ggplot(abund_df, aes(x=group_levels, y=1, 
                                       color=dist_gap, fill=dist_gap)) +
    geom_point(shape=15, size=5)+
    scale_color_manual(values = c("black", "grey30", "grey65", "grey80", "grey95")) +
    scale_fill_manual(values = c("black", "grey30", "grey65", "grey80", "grey95")) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 145), breaks=seq(0, 145, 5),
                       labels = seq(0, 1450, 50)) +
    scale_y_continuous(expand = c(0, 1), limits = NULL, breaks = NULL) +
    labs(x = "Distance from Yichang (km)",
         y = "Porpoise distribution") +
    guides(fill=guide_legend(""), color=guide_legend("")) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")
abund_gap_plot
```

__Plot the segment-wise abundance changes__   

```{r}
pos_df <- abund_df %>%
  dplyr::select(abund_change, group_levels) %>%
  mutate(category = "positive") %>%
  mutate(seg_abund = case_when(abund_change < 0 ~ 0,
                             TRUE ~ abund_change))
  
neg_df <- abund_df %>%
  dplyr::select(abund_change, group_levels) %>%
  mutate(category = "negative") %>%
  mutate(seg_abund = case_when(abund_change >= 0 ~ 0,
                             TRUE ~ abund_change)) 

comb_df <- bind_rows(pos_df, neg_df) %>%
  mutate(category=as.factor(category))

head(comb_df, 5)
```

```{r}
ggplot(comb_df, aes(x=group_levels, y=seg_abund, fill=category))+
    geom_area(linewidth = 1) + 
    scale_fill_manual(values = c("goldenrod3", "skyblue3")) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 145), breaks=seq(0, 145, 5),
                       labels = seq(0, 1450, 50)) +
    labs(x = "Distance from Yichang (km)") +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")

```



## 2. Plot the segment-wise changing trend of the floodplain area  

We grouped river segments into decreased and stable categories given whether 
the floodplain area showed a significant decline. We grouped the nine segments 
with a decreased floodplain area into the stable category (see Appendix S3 for 
more details).  

```{r}
seg_wise_floodplain_df <- read.csv(file="floodplain_area_regression_results.csv")
head(seg_wise_floodplain_df, 5)
```

```{r}
group_lvls <- seg_wise_floodplain_df$groups
floodplain_df <- seg_wise_floodplain_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    mutate(flp_area_trend = case_when((p_value < 0.05 & slope < 0) ~ "decreased",
                                TRUE ~ "stable")) %>%
    mutate(flp_area_trend = factor(flp_area_trend, 
                                    levels = c("decreased", "stable"))) 
head(floodplain_df, 5)
```

```{r}
floodplain_plot <- ggplot(floodplain_df, aes(
        x=group_levels, y=1, color=flp_area_trend, fill=flp_area_trend)) +
    geom_point(shape=22, size=5)+
    scale_color_manual(values = c("black", "grey80"))+
    scale_fill_manual(values = c("black", "grey80")) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 145), breaks=seq(0, 145, 5),
                       labels = seq(0, 1450, 50)) +
    scale_y_continuous(expand = c(0, 1), limits = NULL, breaks = NULL) +
    labs(x = "Distance from Yichang (km)",
         y = "Floodplain area trend") +
    guides(fill=guide_legend(""), color=guide_legend("")) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")

floodplain_plot
```

## 3. Plot the segment-wise changing trend of the cropland%    

We grouped river segments into increased and stable categories given whether 
the cropland% showed a significant increase. We grouped the four segments 
with an increased cropland% into the stable category (see Appendix S3 for 
more details).  

```{r}
seg_wise_cropland_df <- read.csv(file="cropland_perc_regression_results.csv")
head(seg_wise_cropland_df, 5)
```


```{r}
cropland_df <- seg_wise_cropland_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    mutate(crop_perc_trend = case_when((p_value < 0.05 & slope > 0) ~ "increased",
                             TRUE ~ "stable")) %>% 
    mutate(crop_perc_trend = factor(crop_perc_trend, 
                                    levels = c("increased", "stable"))) 
head(cropland_df, 5)
```

```{r}
cropland_plot <- ggplot(cropland_df , aes(
        x=group_levels, y=1, color=crop_perc_trend, fill=crop_perc_trend)) +
    geom_point(shape=22, size=5)+
    scale_color_manual(values = c("black", "grey80"))+
    scale_fill_manual(values = c("black", "grey80")) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 145), breaks=seq(0, 145, 5),
                       labels = seq(0, 1450, 50)) +
    scale_y_continuous(expand = c(0, 1), limits = NULL, breaks = NULL) +
    labs(x = "Distance from Yichang (km)",
         y = "Cropland% trend") +
    guides(fill=guide_legend(""), color=guide_legend("")) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")
cropland_plot 
```
## 4. Plot the segment-wise urban development levels     

We grouped the “developed” and the “highly developed” as one category as 
“developed”, representing those river segments with relatively large urban% 
and large increment in urban% (see Appendix S3 for more details).  

```{r}
seg_wise_urban_df <- read.csv(file="urban_cluster_results.csv")
head(seg_wise_urban_df, 5)
```


```{r}
urban_df <- seg_wise_urban_df %>%
    mutate(groups = factor(groups, levels=group_lvls)) %>%
    mutate(group_levels = as.numeric(groups)) %>%
    filter(group_levels <= 145) %>%
    droplevels() %>%
    mutate(urban_levels = case_when(
       development_levels=="less-developed" ~ "Less developed",
       development_levels=="fast-developing" ~ "Fast developing",
       TRUE ~ "Developed")) %>%
    mutate(urban_levels = factor(urban_levels, 
       levels = c("Developed", "Fast developing", "Less developed")))
head(urban_df, 5)
```


```{r}
urban_plot <- ggplot(urban_df, aes(
        x=group_levels, y=1, color=urban_levels, fill=urban_levels)) +
    geom_point(shape=22, size=5)+
    scale_color_manual(values = c("black", "grey50", "grey80"))+
    scale_fill_manual(values = c("black", "grey50", "grey80")) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 145), breaks=seq(0, 145, 5),
                       labels = seq(0, 1450, 50)) +
    scale_y_continuous(expand = c(0, 1), limits = NULL, breaks = NULL) +
    labs(x = "Distance from Yichang (km)",
         y = "Urban development") +
    guides(fill=guide_legend(""), color=guide_legend("")) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")

urban_plot
```

```{r}
gap_shishou <- c(4: 27)
recover_jjm <- c(28: 53)
gap_wuhan <- c(54: 66)
continue_large <- c(67: 110)
new_gap_tongling <- c(111: 115)
steady_nanjing <- c(116: 133)
new_gap2 <- c(134: 139)
continue_zhenjiang <- c(140: 144)

feature_list <- list(gap_shishou, recover_jjm, gap_wuhan, continue_large,
                  new_gap_tongling, steady_nanjing, new_gap2, continue_zhenjiang)
```


```{r}
library("scales")

flp_perc_list <- list()

for (i in 1: length(feature_list)) {
    flp_perc_list[[i]] <- floodplain_df %>% 
                          filter(group_levels %in% feature_list[[i]]) %>% 
                          pull(flp_area_trend) %>% 
                          summary()
}

flp_perc_df <- as.data.frame(do.call(rbind, flp_perc_list)) %>%
    mutate(decrease_perc = scales::percent(decreased/(decreased + stable)))
flp_perc_df
```

```{r}
crop_perc_list <- list()

for (i in 1: length(feature_list)) {
    crop_perc_list[[i]] <- cropland_df %>% 
                          filter(group_levels %in% feature_list[[i]]) %>% 
                          pull(crop_perc_trend) %>% 
                          summary()
}

crop_perc_df <- as.data.frame(do.call(rbind, crop_perc_list)) %>%
    mutate(increase_perc = scales::percent(increased/(increased + stable)))
crop_perc_df
```


```{r}
urban_perc_list <- list()

for (i in 1: length(feature_list)) {
    urban_perc_list[[i]] <- urban_df %>% 
                          filter(group_levels %in% feature_list[[i]]) %>% 
                          pull(urban_levels) %>% 
                          summary()
}

urban_perc_df <- as.data.frame(do.call(rbind, urban_perc_list))
colnames(urban_perc_df) <- c("developed", "fast_dev", "less_dev")

urban_perc_df <- urban_perc_df %>%
    mutate(developed_perc = scales::percent(developed/(developed + fast_dev + less_dev))) %>%
    mutate(fast_dev_perc = scales::percent(fast_dev/(developed + fast_dev + less_dev))) %>%
    mutate(less_dev_perc = scales::percent(less_dev/(developed + fast_dev + less_dev)))

urban_perc_df
```

