---
title: "DiD_analysis_PAs_effect_on_porpoises"
author: "Yi Han"  
date: "2024-07-05"
output: html_document
---

**Email**: [yi.han.178\@gmail.com](mailto:yi.han.178@gmail.com){.email}\
**Description**: The following code aims to test (1) whether the porpoise abundance was higher and less prone to decline in the PAs than the non-PAs in 2006 and 2017 (Fig 4a & b) and (2) whether the magnitude of abundance decrease is smaller in the PAs (Fig 4c).

```{r include=FALSE}
library(tidyverse)
library(gghalves)
library(MASS)
library(gridExtra)
library(knitr)
library(kableExtra)
```

## 1. Data preparation

(a) Load the file *segment_wise_porpoise_abundance.csv* produced by *analysis_river_lengths_against_abund_ranks.Rmd*\

```{r}
seg_abund_df <- read.csv(file = "../02_percentage_coverage_of_PAs/segment_wise_porpoise_abundance.csv")

group_levels <- seg_abund_df$groups[1:150]
seg_abund <- seg_abund_df %>%
      mutate(reserve = as.factor(reserve),
             reserve_without_nj = as.factor(reserve_without_nj),
             f_year = as.factor(f_year),
             abund_ranks = as.factor(abund_ranks),
             groups = factor(groups, levels=group_levels))
head(seg_abund, 5)
```

(b) Create a new dataframe with each year as a new column and the corresponding segment-wise abundance as inputs.\

```{r}
seg_abund_wide <- seg_abund %>% 
    dplyr::select(groups, f_year, segment_mean, reserve, reserve_without_nj) %>%
    tidyr::pivot_wider(., names_from = f_year, values_from = segment_mean) 
head(seg_abund_wide, 5)
str(seg_abund_wide)
```
(c) Create a new column stores the values of abundance change between 2006 and 2017.  
```{r}
seg_abund_wide <- seg_abund_wide %>%
    rename(abund_06 = "2006", 
           abund_17 = "2017") %>%
    mutate(abund_change = abund_17-abund_06)
head(seg_abund_wide, 5)
```

```{r}
cat("\n----Propoise abundance summary statistics in 2006 and 2017----\n")
abund_06 <- summary(seg_abund_wide$abund_06)
abund_17 <- summary(seg_abund_wide$abund_17)
cat("Summary statistics in 2006: \n")
print(abund_06)
cat("\n")
cat("Summary statistics in 2017: \n")
print(abund_17)
# wilcox.test(seg_abund_wide$abund_06, seg_abund_wide$abund_17, alternative = "g")
# wilcox.test(seg_abund_wide$abund_06, seg_abund_wide$abund_17)
```

### Mann–Whitney U test to examine whether the porpoise abundance in the PAs is higher than that in the non-PAs in 2006 with three threshold

```{r}
thresholds <- c(0.00001, 0.05, 0.1)

for (threshold in thresholds) {
  cat("\n----------- Results for Exclusion Abundance < ", threshold, " porpoise km-1 -----------\n")

  seg_abund_wide_exl <- seg_abund_wide |> 
    filter(!(abund_06 < threshold & abund_17 < threshold))  |> 
    droplevels()

  cat("\n----Number of Segments excluded----\n")
  seg_abund_low <- seg_abund_wide %>%
    filter((abund_06 < threshold & abund_17 < threshold)) 
  print(summary(seg_abund_low$reserve))
  
  abund_06_PA <- seg_abund_wide_exl %>% filter(reserve_without_nj=="Y") %>% pull(abund_06)
  abund_06_nonPA <- seg_abund_wide_exl %>% filter(reserve_without_nj=="N") %>% pull(abund_06)

  outputs_06 <- c(mean(abund_06_PA), sd(abund_06_PA), mean(abund_06_nonPA), sd(abund_06_nonPA))
  names(outputs_06) <- c("abund_06_PA_mean", "abund_06_PA_sd", "abund_06_nonPA_mean", "abund_06_nonPA_sd")

  cat("\n\n----Statistics for 2006 at threshold ", threshold, ":----\n")
  print(outputs_06)
  wilcox_06 <- wilcox.test(abund_06 ~ reserve_without_nj, data=seg_abund_wide_exl)
  cat("\n\n----Mann–Whitney U test for 2006:----\n")
  print(wilcox_06)
  
  abund_17_PA <- seg_abund_wide_exl %>% filter(reserve=="Y") %>% pull(abund_17)
  abund_17_nonPA <- seg_abund_wide_exl %>% filter(reserve=="N") %>% pull(abund_17)
  
  outputs_17 <- c(mean(abund_17_PA), sd(abund_17_PA), mean(abund_17_nonPA), sd(abund_17_nonPA))
  names(outputs_17) <- c("abund_17_PA_mean", "abund_17_PA_sd", "abund_17_nonPA_mean", "abund_17_nonPA_sd")
  
  cat("\n-----Statistics for 2017 at threshold ", threshold, ":----\n")
  print(outputs_17)
  wilcox_17 <- wilcox.test(abund_17 ~ reserve, data=seg_abund_wide_exl)
  cat("\n\n----Mann–Whitney U test for 2017:----\n")
  print(wilcox_17)
}
```

## Difference-in-difference analysis

### Threshold: excluding river segments with 0 porpoise km-1 in both 2006 and 2017.
```{r}
threshold <- 0.000001

seg_abund_wide_exl <- seg_abund_wide |> 
    filter(!(abund_06 < threshold & abund_17 < threshold))  |> 
    droplevels()
# str(seg_abund_wide_exl)
```

```{r}
# seg_abund_wide_exl

seg_abund_wide_exl_long <- seg_abund_wide_exl |>
  dplyr::select(groups, abund_06, abund_17, reserve) |> 
  pivot_longer(
    cols = starts_with("abund"),
    names_to = "year",
    values_to = "abundance"
  ) |> 
  mutate(year = recode(year, abund_06 = "2006", abund_17 = "2017"))

# seg_abund_wide_exl_long
```

```{r}
# Create the binary variables for DID
seg_abund_wide_exl_binary <- seg_abund_wide_exl_long %>%
  mutate(
    PA = ifelse(reserve == "Y", 1, 0),
    Post = ifelse(year == 2017, 1, 0)
  )
head(seg_abund_wide_exl_binary)
```

```{r}
did_model <- lm(abundance ~ PA + Post + PA:Post, data = seg_abund_wide_exl_binary)
summary(did_model)
```

```{r}
library(sandwich)
library(lmtest)
library(zoo)
coeftest(did_model, vcov = vcovHC(did_model, type = "HC1"))

```

```{r}
# Creating a data frame from your model output
results <- data.frame(
  Term = c("Intercept", "PA", "Post", "PA:Post"),
  Estimate = round(c(0.13375, 0.079335, -0.034864, -0.058372), 4),
  `Std Error` = round(c(0.013439, 0.029305, 0.017415, 0.034944), 4),
  `t value` = round(c(9.9520, 2.7072, -2.0020, -1.6704), 4),
  `p-value` = c("< 2.2e-16", "0.0072", "0.0462", "0.0959")
)
results$Significance <- c("***", "**", "*", ".")

kable(results, format = "html", caption = "Statistical Results of Difference-in-Differences Analysis")

results_table <- kable(results, format = "html", booktabs = TRUE, 
                       caption = "Statistical Results of Difference-in-Differences Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>% # Make the first column bold
  column_spec(5, color = "red") # Change the color for p-values for emphasis

# To display the table in R Markdown or output it to an HTML report
results_table
```


### Threshold: excluding river segments with porpoise abundance < 0.05 porpoise km-1 in both 2006 and 2017.
```{r}
threshold <- 0.05

seg_abund_wide_exl <- seg_abund_wide |> 
    filter(!(abund_06 < threshold & abund_17 < threshold))  |> 
    droplevels()
# str(seg_abund_wide_exl)

```

```{r}
# seg_abund_wide_exl

seg_abund_wide_exl_long <- seg_abund_wide_exl |>
  dplyr::select(groups, abund_06, abund_17, reserve) |> 
  pivot_longer(
    cols = starts_with("abund"),
    names_to = "year",
    values_to = "abundance"
  ) |> 
  mutate(year = recode(year, abund_06 = "2006", abund_17 = "2017"))

# seg_abund_wide_exl_long
```

```{r}
# Create the binary variables for DID
seg_abund_wide_exl_binary <- seg_abund_wide_exl_long %>%
  mutate(
    PA = ifelse(reserve == "Y", 1, 0),
    Post = ifelse(year == 2017, 1, 0)
  )
head(seg_abund_wide_exl_binary)
```

```{r}
did_model <- lm(abundance ~ PA + Post + PA:Post, data = seg_abund_wide_exl_binary)
summary(did_model)
```

```{r}
coeftest(did_model, vcov = vcovHC(did_model, type = "HC1"))
```

```{r}
# Creating a data frame from your model output
results <- data.frame(
  Term = c("Intercept", "PA", "Post", "PA:Post"),
  Estimate = round(c(0.157954, 0.088957, -0.041158, -0.068487), 4),
  `Std Error` = round(c(0.014813, 0.031462, 0.019349, 0.037840), 4),
  `t value` = round(c(10.6634, 2.8274, -2.1271, -1.8099), 4),
  `p-value` = c("< 2.2e-16", "0.0051", "0.0344", "0.0716")
)
results$Significance <- c("***", "**", "*", ".")

kable(results, format = "html", caption = "Statistical Results of Difference-in-Differences Analysis")

results_table <- kable(results, format = "html", booktabs = TRUE, 
                       caption = "Statistical Results of Difference-in-Differences Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>% # Make the first column bold
  column_spec(5, color = "red") # Change the color for p-values for emphasis

# To display the table in R Markdown or output it to an HTML report
results_table
```

### Threshold: excluding river segments with porpoise abundance < 0.1 porpoise km-1 in both 2006 and 2017.
```{r}
threshold <- 0.1

seg_abund_wide_exl <- seg_abund_wide |> 
    filter(!(abund_06 < threshold & abund_17 < threshold))  |> 
    droplevels()
# str(seg_abund_wide_exl)

```

```{r}
# seg_abund_wide_exl

seg_abund_wide_exl_long <- seg_abund_wide_exl |>
  dplyr::select(groups, abund_06, abund_17, reserve) |> 
  pivot_longer(
    cols = starts_with("abund"),
    names_to = "year",
    values_to = "abundance"
  ) |> 
  mutate(year = recode(year, abund_06 = "2006", abund_17 = "2017"))

# seg_abund_wide_exl_long
```

```{r}
# Create the binary variables for DID
seg_abund_wide_exl_binary <- seg_abund_wide_exl_long %>%
  mutate(
    PA = ifelse(reserve == "Y", 1, 0),
    Post = ifelse(year == 2017, 1, 0)
  )
head(seg_abund_wide_exl_binary)
```

```{r}
did_model <- lm(abundance ~ PA + Post + PA:Post, data = seg_abund_wide_exl_binary)
summary(did_model)
```

```{r}
coeftest(did_model, vcov = vcovHC(did_model, type = "HC1"))
```

```{r}
# Creating a data frame from your model output
results <- data.frame(
  Term = c("Intercept", "PA", "Post", "PA:Post"),
  Estimate = round(c(0.190178, 0.101795, -0.055263, -0.081872), 4),
  `Std Error` = round(c(0.016708, 0.033190, 0.022473, 0.041282), 4),
  `t value` = round(c(11.3827, 3.0671, -2.4591, -1.9833), 4),
  `p value` = c("< 2.2e-16", "0.0025", "0.0148", "0.0488")
)
results$Significance <- c("***", "**", "*", ".")

kable(results, format = "html", caption = "Statistical Results of Difference-in-Differences Analysis")

results_table <- kable(results, format = "html", booktabs = TRUE, 
                       caption = "Statistical Results of Difference-in-Differences Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>% # Make the first column bold
  column_spec(5, color = "red") # Change the color for p-values for emphasis

# To display the table in R Markdown or output it to an HTML report
results_table
```


### Plotting
__Plot porpoise abundance in 2006 (PAs v.s. non-PAs)__  

```{r}
abund_06 <- ggplot(seg_abund_wide_exl, aes(x = reserve_without_nj, y=abund_06)) +
    geom_half_boxplot(fill="grey70")+
    geom_half_point(color="grey30")+
    scale_y_continuous(limits = c(0, 0.9), breaks=seq(0, 0.9, 0.2)) +
    scale_x_discrete(limits = c("Y", "N"),
                     breaks = c("Y", "N"),
                     labels = c("FPAs", "non-FPAs")) +
    labs (x = "", y = expression(paste("Encounter rate in 2006 ", "(", "porpoise km"^" -1", ")"))) +
    theme_classic(base_size = 14) +
    theme(legend.position.inside = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")
abund_06
```

__Plot porpoise abundance in 2017 (PAs v.s. non-PAs)__  
```{r}
abund_17 <- ggplot(seg_abund_wide_exl, aes(x = reserve, y=abund_17)) +
    geom_half_boxplot(fill="grey70")+
    geom_half_point(color="grey30")+
    scale_y_continuous(limits = c(0, 0.6), breaks=seq(0, 0.6, 0.2)) +
    scale_x_discrete(limits = c("Y", "N"),
                     breaks = c("Y", "N"),
                     labels = c("FPAs", "non-FPAs")) +
    labs (x = "", y = expression(paste("Encounter rate in 2017 ", "(", "porpoise km"^" -1", ")"))) +
    theme_classic(base_size = 14) +
    theme(legend.position = c(.99, .99),
          legend.justification = c("right", "top"),
          legend.box.just = "right")

abund_17
```

__Plot DiD analysis results__  

```{r}
data_summary <- seg_abund_wide_exl_binary %>%
  group_by(PA, year) %>%
  summarise(Abundance = mean(abundance), .groups = 'drop') |> 
  mutate()

data_summary
```


```{r}
data <- data.frame(
  Group = rep(c("FPAs", "non-FPAs"), each = 2),
  Year = rep(c("2006", "2017"), 2),
  Abundance = c(0.2919730, 0.1548378, 0.1901780, 0.1349153)  # Sample mean values
)

abund_change <- ggplot(data, aes(x = Year, y = Abundance, group = Group)) +
  geom_line(aes(linetype = Group), size = 1.2) +
  geom_point(size = 3) +
  # scale_color_manual(values = c("FPA" = "black", "non-FPA" = "black")) +
  labs(x = "Year",
       y = expression(paste("Average encounter rate ", "(", "porpoise km"^" -1", ")"))
       ) +
  theme_classic(base_size = 14) +
  theme(legend.position = c(.99, .99),
        legend.justification = c("right", "top"),
        legend.box.just = "right",
        legend.title = element_blank())
abund_change
```


```{r fig.width=8, fig.height=4, tags=c()}

combined_plot <- grid.arrange(abund_06, abund_17, abund_change, ncol = 3) 
combined_plot
```
```{r}
ggsave("combined_plots.png", combined_plot, width = 12, height = 4)
```

